---
title: 'Gun Violence in the US: Exploratory Data Analysis'
output: html_document
---

Load libraries and the dataset:
```{r load, cache = TRUE, results = FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(RColorBrewer)

df <- read_csv("data/gun-violence-data_01-2013_03-2018.csv")
```

There are a few columns we don't care about. Just keep the ones we want:
```{r}
gun <- df %>% select(incident_id, date, state, city_or_county, address, n_killed, n_injured, gun_stolen, gun_type, incident_characteristics, latitude, location_description, longitude, n_guns_involved, notes, participant_age, participant_age_group, participant_gender, participant_name, participant_relationship, participant_status, participant_type) 
```

So how many gun casualties (defined as number killed and injured) were there for each year?
```{r}
gun %>% group_by(Year = format(date, "%Y")) %>% summarize(n = sum(n_injured + n_killed)) %>% ggplot(aes(x = Year, y = n)) + geom_col()
```

Clearly there are missing data from 2013, and 2018 isn't complete yet. Let's drop these two years. Let's also separate the number of "battery" or "homicide" incidents from suicides. 
```{r}
gun <- filter(gun, format(date, "%Y") != 2018 & format(date, "%Y") != 2013)
suicide <- filter(gun, grepl("\\|\\|Suicide\\^", incident_characteristics)
                      & n_killed + n_injured < 2)

gun <- filter(gun, !(incident_id %in% suicide$incident_id))
```

How many gun casualties were there in each state for 2014-2017?

```{r fig.width=10}
cols <- brewer.pal(3, "Dark2")
gun %>% group_by(state) %>% 
      summarize(n_casualty = sum(n_killed + n_injured)) %>% 
      ggplot(aes(x = reorder(state, -n_casualty), y = n_casualty)) + 
      geom_col(fill = cols[1]) + 
      theme(axis.text.x  = element_text(angle=90, vjust=0.5, size = 14)) + 
      theme(axis.title.y = element_text(size = 16)) + 
      theme(plot.title = element_text(size = 16)) + 
      labs(x = NULL, y = "Casualty (Injured + Killed)", 
           title = "Number of Gun Casualties in the US by State for 2014-2017")
```